---
title: "Lecture 2:  \n Choice of the Target Allocation Ratio in Clinical Trials"
author: 
  - Yevgen Ryeznik <sup>&copy;</sup> (Department of Mathematics, Uppsala University)
  - Oleksandr Sverdlov <sup>&copy;</sup> (Early Development Biostatistics, Novartis Institutes for Biomedical Research)
css: style.css
runtime: shiny
output: 
  slidy_presentation:
    df_print: paged
---

```{r setup, include=FALSE}
library(shiny)
library(tidyverse)
library(kableExtra)

source("./code/lecture02-code.R")
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 8, 
  fig.height = 6, 
  fig.align = "center"
)
```

## Agenda

1. Maximizing power of treatment comparison
2. Maximizing estimation precision of treatment effects and contrasts
3. Optimal allocation designs to balance statistical efficiency and ethical considerations
4. Optimal designs for dose-response studies

## Background

- The design of any cinical trial starts with formulation of study objectives, e.g.:

    + To test whether the effect of an experimental treatment is different from control
    + To estimate treatment contrasts (experimental group vs. control)
    + To minimize the number of treatment failures while maintaining study power
    + To estimate dose-response relationship
    + Etc. ...

- Once study objectives are quantified, one may want to obtain an optimal design to achive the desired objectives

## Statistical methodologies

```{r methodologies, echo=FALSE}
tab <- tibble::tibble(
  ` ` = c('Research Question(s)', 'Statistical Tool', 'Criteria for "goodness"'),
  `Hypothesis Testing` = c('Is experimental treatment better than control? (Yes/No)',
                           'Test of significance (e.g. two-sample t-test)',
                           '- Statistical test should have **high power** to detect
statisticaly significant difference between groups\n - **Validity** -- Pr(type I error) should be maintained at the nominal level (e.g. 5%)'),
  `Estimation` = c('- What is the mean difference between experimental and control?\n - What is the estimated dose-response curve?',
                   'Estimator', 'Estimators should be **unbiased** and have **low variance**')
)
knitr::kable(tab, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(c(1,3), background = "#99cfe0") %>% 
  row_spec(2, background = "#c1e1ec")
```


## I. Maximizing power of a 2-arm comparative trial

- Two groups: Experimental (E) and Control (C)
- Sample size: a total of $n = n_E+n_C$ subjects

    + $n_E = \rho n$ assigned to E, and $n_C = (1-\rho)n$ assigned to C (if necessary, rounded to the integer values)
    + $\rho \in (0, 1)$ -- "allocation ratio"
    
- Primary outcomes: $Y_{ik}\sim Normal(\mu_k, \sigma^2)$, $k=E,C$; $i = 1, 2, \ldots, n_k$

    + **$\sigma$ is common for both groups, assumed known**

- Study objective: to test $H_0: \mu_E = \mu_C$ vs. $H_1: \mu_E > \mu_C$ (experimental is better than control)
- $Z$-test: reject $H_0$ at level $\alpha$, if $Z = \frac{\overline{Y}_E-\overline{Y}_C}{\sigma\sqrt{\frac{1}{n_E}+\frac{1}{n_C}}} > z_{1-\alpha}$ 

    + (e.g. $\alpha=0.05 \Rightarrow$ reject $H_0$ at 5% significance level, if $Z > 1.645$)


## Which allocation maximizes power? (1 of 2)

- _**Power**_ = 
$\begin{aligned}
 &= \text{Pr}(\text{Reject} H_0|n_E, n_C, \mu_E, \mu_C, \sigma^2, z_{1-\alpha}) = \text{Pr}\left(\frac{\overline{Y}_E-\overline{Y}_C}{\sigma\sqrt{\frac{1}{n_E}+\frac{1}{n_C}}} > z_{1-\alpha}\right) = \\
&= \left| \text{Let } \left\{
\begin{array}{l}
  Z_0\sim N(0, 1); \\
  \Phi\left(\cdot\right) - \text{ c.d.f. of standard normal R.V.}; \\
  \Delta = \frac{\mu_E-\mu_C}{\sigma\sqrt{\frac{1}{n_E}+\frac{1}{n_C}}} > 0 \text{ (because }\mu_E>\mu_C\text{ under }H_1\text{)}
\end{array}\right.
\right| = \\
&= \text{Pr}\left(\frac{\overline{Y}_E-\overline{Y}_C - (\mu_E-\mu_C)}{\sigma\sqrt{\frac{1}{n_E}+\frac{1}{n_C}}} + \frac{(\mu_E-\mu_C)}{\sigma\sqrt{\frac{1}{n_E}+\frac{1}{n_C}}} > z_{1-\alpha}\right) = \\
&= \text{Pr}\left(Z_0+\Delta\right) = 1-\Phi\left(z_{1-\alpha}-\Delta\right)
\end{aligned}$

- To maximize _**Power**_, we need to minimize $\Phi\left(z_{1-\alpha}-\Delta\right)$ $\Leftrightarrow$ to maximize $\Delta$ w.r.t. $\rho \in (0,1)$


## Which allocation maximizes power? (2 of 2)

- $\Delta = \frac{\mu_E-\mu_C}{\sigma\sqrt{\frac{1}{n\rho}+\frac{1}{n(1-\rho)}}}\rightarrow \max\limits_{\rho \in (0,1)}$
- Equivalently, $\frac{1}{n\rho}+\frac{1}{n(1-\rho)}\rightarrow \min\limits_{\rho\in (0,1)}$
- Optimal solution is $\boldsymbol{\rho^* = \frac{1}{2}}$ (_**equal allocation**_)

<span style="background-color: #6666ff; color:white">
Therefore, for a 2-arm comparative trial with normal outcomes (with known variance, common to 2 groups), equal allocation $\left(n_E=n_c=\frac{n}{2}\right)$ maximizes the power of $Z$-test for treatment comparison.
</span>

**Note**: Same result is obtained, if we consider 2-sample t-test instead of $Z$-test.


## Which allocation maximizes power? -- Demo 

```{r, echo = FALSE}
shinyApp(
  
  ui = fluidPage(
    sidebarPanel(
      withMathJax(),
      sliderInput("sample_size", 
                  "$$\\text{sample size }(n):$$", 
                  min = 10, 
                  max = 30,
                  step = 2,
                  value = 10),
      sliderInput("means_difference", 
                  "$$\\text{means difference }(\\mu_E-\\mu_C):$$", 
                  min = 0, 
                  max = 2,
                  step = 0.1, 
                  value = 0.1),
      sliderInput("stddev", 
                  "$$\\text{standard deviation }(\\sigma):$$", 
                  min = 1, 
                  max = 2.5,
                  step = 0.1, 
                  value = 1),
      sliderInput("alpha", 
                  "$$\\text{significance level }(\\alpha)$$", 
                  min = 0.01, 
                  max = 0.2,
                  step = 0.01, 
                  value = 0.05)


    ),
    mainPanel(
      plotOutput("power_plot")
    )
  ),
  
  server = function(input, output) {
    output$power_plot <- renderPlot({
      rho <- seq(0.1, 0.9, by = 0.1)
      
      means_difference <- as.numeric(input$means_difference)
      sigma <- as.numeric(input$stddev)
      alpha <- as.numeric(input$alpha)
      z <- qnorm(1-alpha)
      
      nE <- as.numeric(input$sample_size)*rho
      nC <- as.numeric(input$sample_size)*(1-rho)
      Delta <- means_difference/(sigma*sqrt(1/nE+1/nC))
        
      pow <- 1-pnorm(z-Delta)
      power_df <- tibble::tibble(rho, pow) 
        ggplot()+
          geom_point(data = power_df, aes(x = rho, y = pow), size = 1.5)+
          geom_line(data = power_df, aes(x = rho, y = pow), size = 1.25)+
          geom_line(data = tibble::tibble(x = rep(0.5, 2), y = c(0, 1)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25)+
          xlab(expression(rho))+
          ylab("Power")+
          scale_y_continuous(limits = c(0, 1), 
                             breaks = seq(0, 1, by = 0.2))+
          scale_x_continuous(breaks = seq(0.1, 0.9, by = 0.1))
    })
  },
  
  options = list(height = 500)
)
     
```

## What if group variances are unequal? (1 of 2)

- $Y_{ik}\sim Normal(\mu_k, \sigma_k^2)$; $k = E, C$, $i = 1, 2, \ldots, n_k$

    + **$\sigma_k$ assumed to be known**, but depends on the treatment group

- To test $H_0: \mu_E = \mu_C$ vs. $H_1: \mu_E > \mu_C$, use $Z$-test: Reject $H_0$ at level $\alpha$, if $Z=\frac{\overline{Y}_E-\overline{Y}_C}{\sqrt{\frac{\sigma^2_E}{n_E}+\frac{\sigma^2_C}{n_C}}} > z_{1-\alpha}$
- As before, _**Power**_ = $\text{Pr}\left(\text{Reject } H_0| n_E, n_C, \mu_E, \mu_C, \sigma_E, \sigma_C, z_{1-\alpha}\right) = 1-\Phi\left(z_{1-\alpha}-\Delta\right)$, where $\Delta = \frac{\mu_E-\mu_C}{\sqrt{\frac{\sigma^2_E}{n_E}+\frac{\sigma^2_C}{n_C}}}$
- To maximize _**Power**_, we need to minimize $\Phi\left(z_{1-\alpha}-\Delta\right)$ $\Leftrightarrow$ to maximize $\Delta$ w.r.t. $\rho \in (0,1)$

## What if group variances are unequal? (2 of 2)

- $\Delta = \frac{\mu_E-\mu_C}{\sqrt{\frac{\sigma^2_1}{n\rho}+\frac{\sigma^2_2}{n(1-\rho)}}}\rightarrow \max\limits_{\rho \in (0,1)}$
- Equivalently, $\frac{1}{n}\left(\frac{\sigma^2_1}{\rho}+\frac{\sigma^2_2}{(1-\rho)}\right)\rightarrow \min\limits_{\rho\in (0,1)}$
- Optimal solution is $\boldsymbol{\rho_N^* = \frac{\sigma_E}{\sigma_E+\sigma_C}}$ ([_**Neyman allocation**_](https://en.wikipedia.org/wiki/Jerzy_Neyman))

<span style="background-color: #6666ff; color:white">
More variable treatment group receives greater proportion of study patients.
</span>
 
 
## What if group variances are unequal? -- Demo 

```{r, echo = FALSE}
shinyApp(
  
  ui = fluidPage(
    sidebarPanel(
      withMathJax(),
      sliderInput("sample_size", 
                  "$$\\text{sample size }(n):$$", 
                  min = 10, 
                  max = 30,
                  step = 2,
                  value = 10),
      sliderInput("means_difference", 
                  "$$\\text{means difference }(\\mu_E-\\mu_C):$$", 
                  min = 0, 
                  max = 2,
                  step = 0.1, 
                  value = 0.1),
      sliderInput("stddevE", 
                  "$$\\text{standard deviation of group E }(\\sigma_E):$$", 
                  min = 1, 
                  max = 2.5,
                  step = 0.1, 
                  value = 1),
      sliderInput("stddevC", 
                  "$$\\text{standard deviation of group C }(\\sigma_C):$$", 
                  min = 1, 
                  max = 2.5,
                  step = 0.1, 
                  value = 1),
      sliderInput("alpha", 
                  "$$\\text{significance level }(\\alpha)$$", 
                  min = 0.01, 
                  max = 0.2,
                  step = 0.01, 
                  value = 0.05)
    ),
    mainPanel(
      plotOutput("power_plot")
    )
  ),
  
  server = function(input, output) {
    output$power_plot <- renderPlot({
      rho <- seq(0.1, 0.9, by = 0.1)
      
      means_difference <- as.numeric(input$means_difference)
      sigmaE <- as.numeric(input$stddevE)
      sigmaC <- as.numeric(input$stddevC)
      rho_star <- sigmaE/(sigmaE+sigmaC)
      alpha <- as.numeric(input$alpha)
      z <- qnorm(1-alpha)
      
      nE <- as.numeric(input$sample_size)*rho
      nC <- as.numeric(input$sample_size)*(1-rho)
      Delta <- means_difference/(sqrt(sigmaE^2/nE+sigmaC^2/nC))
        
      pow <- 1-pnorm(z-Delta)
      power_df <- tibble::tibble(rho, pow) 
        ggplot()+
          geom_point(data = power_df, aes(x = rho, y = pow), size = 1.5)+
          geom_line(data = power_df, aes(x = rho, y = pow), size = 1.25)+
          geom_line(data = tibble::tibble(x = rep(rho_star, 2), y = c(0, 1)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25)+
          xlab(expression(rho))+
          ylab("Power")+
          scale_y_continuous(limits = c(0, 1), 
                             breaks = seq(0, 1, by = 0.2))+
          scale_x_continuous(breaks = seq(0.1, 0.9, by = 0.1))
    })
  },
  
  options = list(height = 500)
)
     
```

## Example: Binary outcome trial

- $Y_{ik}\sim Bernoulli(p_k)$; $k = E, C$; $i = 1, 2, \ldots, n_k$

    + $p_k$ = probability of a successful outcome in group $k$
    + Denote $q_k = 1-p_k$ = probability of a "failure" in group $k$
    
- We want to test $H_0: p_E = p_C$ vs. $p_E > p_C$ by using (asymptotic) $Z$-test:

    + Reject $H_0$ at level $\alpha$, if $Z=\frac{\widehat{p}_E-\widehat{p}_C}{\sqrt{\frac{\widehat{p}_E\widehat{q}_E}{n_E}+\frac{\widehat{p}_C\widehat{q}_C}{n_C}}} > z_{1-\alpha}$ (where $\widehat{p}_k, \widehat{q}_k$ are sample proportions)
    + Asymptotically $\widehat{p}_k \rightarrow p_k$, $\widehat{p}_k\widehat{q}_k \rightarrow p_kq_k$, all our derivations (developed in the normal outcome case) still apply here
    
- Note that $\text{E}\left[Y_{ik}\right]=p_k$ and $\text{Var}\left[Y_{ik}\right]=p_kq_k$ (variances are possibly different in two groups)
$$\rho_N^* = \frac{\sqrt{p_Eq_E}}{\sqrt{p_Eq_E}+\sqrt{p_Cq_C}}$$  (_**Neyman allocation**_)

## Example: Binary outcome trial -- Demo 

```{r, echo = FALSE}
shinyApp(
  ui = fluidPage(
    sidebarPanel(
      withMathJax(),
      sliderInput("sample_size", 
                  "$$\\text{sample size }(n):$$", 
                  min = 50, 
                  max = 100,
                  step = 5,
                  value = 100),
      sliderInput("probE", 
                  "$$\\text{probability of success in group E }(p_E):$$", 
                  min = 0.05, 
                  max = 0.95,
                  step = 0.05, 
                  value = 0.05),
      sliderInput("probC", 
                  "$$\\text{probability of success in group C }(p_C):$$", 
                  min = 0.05, 
                  max = 0.95,
                  step = 0.05, 
                  value = 0.05),
      sliderInput("alpha", 
                  "$$\\text{significance level }(\\alpha)$$", 
                  min = 0.01, 
                  max = 0.2,
                  step = 0.01, 
                  value = 0.05)
    ),
    mainPanel(
      plotOutput("power_plot")
    )
  ),
  
  server = function(input, output) {
    output$power_plot <- renderPlot({
      rho <- seq(0.1, 0.9, by = 0.1)
      
      probE <- as.numeric(input$probE)
      probC <- as.numeric(input$probC)
      rho_star <- sqrt(probE*(1-probE))/(sqrt(probE*(1-probE))+sqrt(probC*(1-probC)))
      alpha <- as.numeric(input$alpha)
      z <- qnorm(1-alpha)
      
      nE <- as.numeric(input$sample_size)*rho
      nC <- as.numeric(input$sample_size)*(1-rho)
      Delta <- (probE-probC)/(sqrt(probE*(1-probE)/nE+probC*(1-probC)/nC))
        
      pow <- 1-pnorm(z-Delta)
      power_df <- tibble::tibble(rho, pow) 
        ggplot()+
          geom_point(data = power_df, aes(x = rho, y = pow), size = 1.5)+
          geom_line(data = power_df, aes(x = rho, y = pow), size = 1.25)+
          geom_line(data = tibble::tibble(x = rep(rho_star, 2), y = c(0, 1)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25, color="blue")+
          geom_line(data = tibble::tibble(x = rep(0.5, 2), y = c(0, 1)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25, color="red")+
          geom_text(aes(x=rho_star, y = 1, label = "Neyman"))+
          xlab(expression(rho))+
          ylab("Power")+
          scale_y_continuous(limits = c(0, 1), 
                             breaks = seq(0, 1, by = 0.2))+
          scale_x_continuous(breaks = seq(0.1, 0.9, by = 0.1))
    })
  },
  
  options = list(height = 500)
)
     
```

***

**Q**: What is a major limitation of Neyman allocation $\rho^* = \frac{\sqrt{p_Eq_E}}{\sqrt{p_Eq_E}+\sqrt{p_Cq_C}}$?

**A**: The model parameters _**are unknown**_ to us! We are conducting the trial to gain information about $p_E$ and $p_C$. Thus Neyman allocation cannot be implemented directly!


## II. Maximizing estimation precision

- $K \geq 2$ treatment groups (e.g. different doses of a drug or different intervention strategies)
- Primary outcome: $Y_{ik}$ follows some distribution from an exponential family (e.g. Normal, Bernoulli, Poisson, exponential, etc.)
- $\mathbf{E}\left[Y_{ik}\right] = \mu_k$, $\mathbf{Var}\left[Y_{ik}\right] = \sigma^2_k$, $k = 1, 2, \ldots, K$
- Study objectives:

    + The vector of mean treatment effects $\boldsymbol{\mu} = (\mu_1, \mu_2, \ldots, \mu_K)$
    + The vector of treatment contrasts $\boldsymbol{\mu}_C = (\mu_2-\mu_1, \ldots, \mu_K-\mu_1)$
    + Etc ...


## Design Optimization (1 of 2)

- Allocation design: $\boldsymbol{\rho}=(\rho_1, \rho_2, \ldots, \rho_K)$, $0 \leq \rho_k \leq 1$ and $\sum_{k=1}^K{\rho_k} = 1$

    + $\rho_k$ = target allocation proportion for the $k$-th treatment group
    + $n\rho_k$ = sample size of the $k$-th group (integer-rounded, if necessary)
 
- Design Fisher Information Matrix: $$\boldsymbol{M}(\boldsymbol{\rho}) = n\times diag\left\{\frac{\rho_1}{\sigma^2_1}, \frac{\rho_2}{\sigma^2_2}, \ldots, \frac{\rho_K}{\sigma^2_K}\right\}$$

- Variance-covariance matrix of a consistent estimator (MLE) $\widehat{\boldsymbol{\mu}}=\left(\widehat{\mu}_1, \widehat{\mu}_2, \ldots, \widehat{\mu}_K\right)$ of $\boldsymbol{\mu}=\left(\mu_1, \mu_2, \ldots, \mu_K\right)$:

$$\mathbf{Var}\left[\widehat{\boldsymbol{\mu}}\right]=\boldsymbol{M}^{-1}(\boldsymbol{\rho}) = \frac{1}{n}diag\left\{\frac{\sigma^2_1}{\rho_1}, \frac{\sigma^2_2}{\rho_2}, \ldots, \frac{\sigma^2_K}{\rho_K} \right\}$$

- Variance-covariance matrix of a consistent estimator of the contrast vector $\boldsymbol{\mu}_C = \boldsymbol{A}^T\boldsymbol{\mu}=(\mu_2-\mu_1, \ldots, \mu_K-\mu_1)$ 

$$\mathbf{Var}\left[\boldsymbol{A}^T\widehat{\boldsymbol{\mu}}\right]=\boldsymbol{A}^T\boldsymbol{M}^{-1}(\boldsymbol{\rho})\boldsymbol{A} = \frac{1}{n}\left(diag\left\{\frac{\sigma^2_2}{\rho_2}, \ldots, \frac{\sigma^2_K}{\rho_K} \right\}+\frac{\sigma^2_1}{\rho_1}\mathbf{1}\mathbf{1}^T\right),$$
where $\boldsymbol{A}$ is a contrast matrix.


## Design Optimization (2 of 2)

- We are interested in minimizing "variability" of an estimated parameter of interest
- An optimal design problem is to find an allocation $\boldsymbol{\rho}=\left(\rho_1, \rho_2,  \ldots, \rho_K\right)$ that minimizes some criterion of $\boldsymbol{M}^{-1}(\boldsymbol{\rho})$ (or $\boldsymbol{A}^T\boldsymbol{M}^{-1}(\boldsymbol{\rho})\boldsymbol{A}$) subject to constraints $0 \leq \rho_k \leq 1$ and $\sum_{k=1}^K{\rho_k}=1$.

    + _**D-optimal**_: $\text{det}\left\{\boldsymbol{M}^{-1}(\boldsymbol{\rho})\right\} \rightarrow \min\limits_{\boldsymbol{\rho}}$ (volume of the confidence ellipsoid for $\boldsymbol{\mu}$)
    + _**A-optimal**_: $\text{trace}\left\{\boldsymbol{M}^{-1}(\boldsymbol{\rho})\right\} \rightarrow \min\limits_{\boldsymbol{\rho}}$ (sum of the lengths of the major axes in the confidence ellipsoid for $\boldsymbol{\mu}$)
    + _**D<sub>A</sub>-optimal**_: $\text{det}\left\{\boldsymbol{A}^T\boldsymbol{M}^{-1}(\boldsymbol{\rho})\boldsymbol{A}\right\} \rightarrow \min\limits_{\boldsymbol{\rho}}$ (volume of the confidence ellipsoid for contrast vector $\boldsymbol{\mu}_C$)
    + _**A<sub>A</sub>-optimal**_: $\text{trace}\left\{\boldsymbol{A}^T\boldsymbol{M}^{-1}(\boldsymbol{\rho})\boldsymbol{A}\right\} \rightarrow \min\limits_{\boldsymbol{\rho}}$ (sum of the lengths of the major axes in the confidence ellipsoid for contrast vector $\boldsymbol{\mu}_C$)

- Some of these problems can be solved _**analytically**_ (e.g. by using Lagrange multiplier minimization)


## D-Optimal Allocation

$$\text{det}\left\{\boldsymbol{M}^{-1}(\boldsymbol{\rho})\right\}=\frac{\sigma^2_1}{\rho_1}\cdot\frac{\sigma^2_2}{\rho_2}\cdot\ldots\cdot\frac{\sigma^2_K}{\rho_K}\rightarrow\min\limits_{\boldsymbol{\rho}}$$

- Equivalently: $\rho_1\cdot\rho_2\cdot\ldots\cdot\rho_K\rightarrow\max\limits_{\boldsymbol{\rho}}$

- By the geometric-arithmetic mean inequality: 
$$\left(\rho_1\cdot\rho_2\cdot\ldots\cdot\rho_K\right)^\frac{1}{K}\leq\frac{1}{K}\sum_{k=1}^K{\rho_k}=\frac{1}{K}\Rightarrow \rho_1\cdot\rho_2\cdot\ldots\cdot\rho_K \leq \left(\frac{1}{K}\right)^K$$

<span style="background-color: #6666ff; color:white">
Thus, equal allocation design $\boldsymbol{\rho}=\left(1/K, \ldots, 1/K\right)$ is D-optimal.
</span>


## A-Optimal Allocation

$$\text{trace}\left\{\boldsymbol{M}^{-1}(\boldsymbol{\rho})\right\}=\frac{\sigma^2_1}{\rho_1}+\frac{\sigma^2_2}{\rho_2}+\ldots+\frac{\sigma^2_K}{\rho_K}\rightarrow\min\limits_{\boldsymbol{\rho}}$$

- Lagrange multiplier minimization: 
$$\frac{\partial}{\partial{\rho_k}}\left\{\sum_{k=1}^K{\frac{\sigma^2_k}{\rho_k}+\lambda\left(\sum_{k=1}^K{\rho_k}-1\right)}\right\}=-\frac{\sigma^2_k}{\rho_k^2}+\lambda,\quad k = 1, 2, \ldots, K$$
- Setting the derivatives to zero, and using $\sum_{k=1}^K{\rho_k} = 1$, we find the A-optimal allocation as $$\rho^{(A)}_k=\frac{\sigma_k}{\sum_{i=1}^K{\sigma_i}}, \quad k = 1, 2, \ldots, K$$

- If $K = 2$, optimal solution is the Neyman allocation!

- If $\sigma^2_k$'s are all equal, optimal solution is the equal allocation!


## D<sub>A</sub>-Optimal Allocation

$$\text{det}\left\{\boldsymbol{A}^T\boldsymbol{M}^{-1}(\boldsymbol{\rho})\boldsymbol{A}\right\} =\text{det}\left\{diag\left\{\frac{\sigma^2_2}{\rho_2}, \ldots, \frac{\sigma^2_K}{\rho_K} \right\}+\frac{\sigma^2_1}{\rho_1}\mathbf{1}\mathbf{1}^T\right\}\rightarrow\min\limits_{\boldsymbol{\rho}}$$

- Optimal solution is not in closed form

- Wong and Zhu (2008)<sup>1</sup> showed that D<sub>A</sub>-optimal proportions satisfy the following (nonlinear) system of equations:

$$\frac{1}{\rho^*_k}-\frac{1/\sigma^2_k}{\sum_{i=1}^K{\rho^*_i/\sigma^2_i}}=0, \quad k =1, 2, \ldots, K.$$

- If $\sigma^2_k$'s are all equal, the D<sub>A</sub>-optimal solution is the equal allocation!

<br />
<br />
<br />
<br />

<sup>1</sup> Wong, W. K., and Zhu, W.  _"Optimum treatment allocation rules under a
variance heterogeneity model"_. Stat. Med., 27, 4581–4595 (2008).


## D<sub>A</sub>-Optimal Allocation -- Some Properties

i. When $K = 2$, the D<sub>A</sub>-optimal allocation is Neyman allocation: 
$$\rho^*_k = \frac{\sigma_k}{\sigma_1+\sigma_2}, \quad k = 1, 2$$

ii. The D<sub>A</sub>-optimal proportions satisfy: 
$$0 \leq \rho^*_k \leq \frac{1}{K-1}, \quad k = 1, 2, \ldots, K$$

iii. If $\sigma^2_k\rightarrow 0$ and $\sigma^2_k/\sigma^2_j\rightarrow 0$ for all $j \ne k$ (variance in the $k$-th group is small relative to all other group variances), then $\rho^*_k\rightarrow 0$

iv. Let the $K$ treatment groups be such that $\sigma_1\geq\sigma_2\geq \ldots\sigma_K$. Then the D<sub>A</sub>-optimal proportions satisfy: $\rho^*_1\geq\rho^*_2\geq \ldots\rho^*_K$ (more variable treatment groups receive greater proportions of study subjects)


## D<sub>A</sub>-Optimal Allocation -- Example

- Binary outcome trial, $K=3$ treatment groups
- $p_k$ probability of a successful outcome in group $k = 1, 2, 3$

```{r binary-3-trt-example, echo=FALSE}
prob_tab <- tribble(
  ~id, ~`$p_1$`, ~`$p_2$`, ~`$p_3$`,
  1, 0.10, 0.02, 0.02,
  2, 0.10, 0.30, 0.50,
  3, 0.30, 0.40, 0.60,
  4, 0.40, 0.60, 0.80,
  5, 0.70, 0.30, 0.80,
  6, 0.90, 0.95, 0.99
)

# calculate D-A-optimal proportions
prop_tab <- map(seq_len(nrow(prob_tab)), ~ {
  id <- as.numeric(prob_tab[., 1])
  prob <- as.numeric(prob_tab[.,])[-1]
  sigma_vec <- sqrt(prob*(1-prob))
  prop <- round(daoptimal(sigma_vec), 3)
  tribble(
    ~id, ~`$\\rho^*_1$`, ~`$\\rho^*_2$`, ~`$\\rho^*_3$`,
    id, prop[1], prop[2], prop[3])
}) %>% 
  bind_rows()

# join propbabilities and proportions into a single table
inner_join(
  prob_tab, 
  prop_tab,
  by = "id"
) %>% 
  mutate(
    `$\\rho^*_1$`= cell_spec(`$\\rho^*_1$`, "html",
                             color = if_else(id > 3, "red", "black")),
    `$\\rho^*_2$`= cell_spec(`$\\rho^*_2$`, "html", 
                             color = if_else(id > 3, "red", "black")),
    `$\\rho^*_3$`= cell_spec(`$\\rho^*_3$`, "html",
                             color = if_else(id > 3, "red", "black"))
    ) %>% 
  select(-id) %>% 
  knitr::kable("html", escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c("Treatment success probabilities" = 3,
                     "$D_A$-optimal allocation proportions" = 3)) %>% 
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(c(1, 3, 5), background = "#99cfe0") %>% 
  row_spec(c(2, 4, 6), background = "#c1e1ec") %>% 
  column_spec(c(1, 2, 3), color = "black")
  #row_spec(-1, background = "#c1e1ec") 
```

<span style="background-color: #6666ff; color:white">
Q: Based on these results, what is one caveat of the D<sub>A</sub>-optimal allocation?
</span>

<span style="background-color: #6666ff; color:white">
A: It can assign greater proportions to less successful treatments (cf. last three examples)
</span>


## A<sub>A</sub>-Optimal Allocation    

$$\text{trace}\left\{\boldsymbol{A}^T\boldsymbol{M}^{-1}(\boldsymbol{\rho})\boldsymbol{A}\right\}=\frac{\sigma^2_2}{\rho_2}+\frac{\sigma^2_3}{\rho_3}+\ldots+\frac{\sigma^2_K}{\rho_K}+(K-1)\frac{\sigma^2_1}{\rho_1}\rightarrow\min\limits_{\boldsymbol{\rho}}$$

- Lagrange multiplier minimization: 

$$\frac{\partial}{\partial{\rho_k}}\left\{\sum_{k=2}^K{\frac{\sigma^2_k}{\rho_k}+(K-1)\frac{\sigma^2_1}{\rho_1}+\lambda\left(\sum_{k=1}^K{\rho_k}-1\right)}\right\}=\left\{
\begin{array}{cl}
  -(K-1)\frac{\sigma^2_1}{\rho_1^2}+\lambda, & k = 1 \\
  -\frac{\sigma^2_k}{\rho_k^2}+\lambda, & k = 2, 3, \ldots, K
\end{array}
\right.$$


- Setting the derivatives to zero, and using $\sum_{k=1}^K{\rho_k} = 1$, we find the A<sub>A</sub>-optimal allocation as 
$$
\rho^{(A)}_k=\left\{
\begin{array}{cl}
  \frac{\sigma_1\sqrt{K-1}}{\sigma_1\sqrt{K-1}+\sum_{i=2}^K{\sigma_i}}, & k = 1\\
  \frac{\sigma_k}{\sigma_1\sqrt{K-1}+\sum_{i=2}^K{\sigma_i}}, & k = 2, 3, \ldots, K
\end{array}
\right.
$$

- If $K = 2$, optimal solution is the Neyman allocation!

## Summary of Optimal Allocation Ratios$^*$

<br />

```{r summary-optimal-allocation, echo=FALSE}
tribble(
  ~" ", ~"Equal variances ($K\\geq 2$)", ~"Unequal variances ($K = 2$)", ~"Unequal variances ($K$ &gt; $2$)",  
  "D-optimal", "$1:1:\\ldots:1$", "$1:1$", "$1:1:\\ldots:1$",
  "A-optimal", "$1:1:\\ldots:1$", "$\\sigma_1:\\sigma_2$", "$\\sigma_1:\\sigma_2:\\ldots:\\sigma_K$",
  "D<sub>A</sub>-optimal", "$1:1:\\ldots:1$","$\\sigma_1:\\sigma_2$", "No closed form  solution;must be found numerically",
  "A<sub>A</sub>-optimal", "$\\sqrt{K-1}:1:\\ldots:1$", "$\\sigma_1:\\sigma_2$", "$\\sigma_1\\sqrt{K-1}:\\sigma_2:\\ldots:\\sigma_K$"
) %>% 
  knitr::kable("html", escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(c(1,3), background = "#99cfe0") %>% 
  row_spec(c(2,4), background = "#c1e1ec") %>% 
  column_spec(c(1, 2), color = "black")
  #row_spec(-1, background = "#c1e1ec") 
```

<br />
<br />
<br />
<br />

<sup>*</sup> An allocation ratio $C_1:C_2:\ldots:C_K$ corresponds to the vector pf allocation proportions $\boldsymbol{\rho}=\left(\frac{C_1}{\sum_{k=1}^K{C_k}}, \frac{C_2}{\sum_{k=1}^K{C_k}}, \ldots, \frac{C_K}{\sum_{k=1}^K{C_k}}\right)$


## Relative Efficiency

- Suppose we want to compare two allocation designs: $\boldsymbol{\rho}^{(1)}$ and $\boldsymbol{\rho}^{(2)}$ w.r.t. a given design criterion
- The relative efficiencies (of $\boldsymbol{\rho}^{(1)}$ vs. $\boldsymbol{\rho}^{(2)}$) are defined as follows:

```{r relative-efficiency, echo=FALSE}
tribble(
  ~"Relative D-efficiency", ~"Relative A-efficiency", ~"Relative D<sub>A</sub>-efficiency", ~"Relative A<sub>A</sub>-efficiency",  
  "$\\left(\\frac{\\text{det}\\{\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(2)})\\}}{\\text{det}\\{\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(1)})\\}}\\right)^{1/K}$", 
  "$\\frac{\\text{trace}\\{\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(2)})\\}}{\\text{trace}\\{\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(1)})\\}}$", 
  "$\\left(\\frac{\\text{det}\\{\\boldsymbol{A}^T\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(2)})\\boldsymbol{A}\\}}{\\text{det}\\{\\boldsymbol{A}^T\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(1)})\\boldsymbol{A}\\}}\\right)^{1/\\text{rank}(A)}$", 
  "$\\frac{\\text{trace}\\{\\boldsymbol{A}^T\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(2)})\\boldsymbol{A}\\}}{\\text{trace}\\{\\boldsymbol{A}^T\\boldsymbol{M}^{-1}(\\boldsymbol{\\rho}^{(1)})\\boldsymbol{A}\\}}$"
) %>% 
  knitr::kable("html", escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(1, background = "#99cfe0") %>% 
  column_spec(c(1, 2, 3, 4), color = "black")
  #row_spec(-1, background = "#c1e1ec") 
```

- An allocation that is highly efficient under one criterion may have poor performance under other criteria

    + Therefore, clear understanding of study objectives is essential!
    
## Example -- Efficiency of Equal Allocation

- Binary outcome trial, $K=3$ treatment groups

```{r efficiency-of-equal-allocation, fig.cap = "Relative efficiencies heatmap plots"}
scenario <- list(
  allocation_fcn = list(
    doptimal,
    aoptimal,
    daoptimal,
    aaoptimal
  ),
  efficiency_fcn = list(
    defficiency,
    aefficiency,
    daefficiency,
    aaefficiency
  ),
  label = list(
    "D-efficiency",
    "A-efficiency",
    "D[A]-efficiency",
    "A[A]-efficiency"
  )
) %>% 
  transpose()

p1 <- seq(0.05, 0.95, by=0.05)
p2 <- seq(0.05, 0.95, by=0.05)

efficiency <- function(allocation_fcn, efficiency_fcn, p1, p2) {
  crossing(p1, p2, p3 = c(0.2, 0.5, 0.9)) %>% 
    transpose() %>% 
    map(~ {
      p <- c(.$p1, .$p2, .$p3)
      sigma_vec <- sqrt(p*(1-p))
      rho1 <- rep(1, 3)/3
      rho2 <- allocation_fcn(sigma_vec)
      tibble::tibble(
        efficiency = efficiency_fcn(rho1, rho2, sigma_vec), 
        p1 = p[1],
        p2 = p[2],
        p3 = p[3]
      )
    }) %>% 
    bind_rows()
}

eff <- scenario %>% 
  map(~ {
    allocation_fcn <- .$allocation_fcn
    efficiency_fcn <- .$efficiency_fcn
    label <- .$label
    efficiency(allocation_fcn, efficiency_fcn, p1, p2)  %>% 
      add_column(label)
  }) %>% 
  bind_rows() 

eff %>% 
  mutate(probability = paste0("(p[1]~~p[2]~~", p3, ")"), efficiency = 1/efficiency) %>%
  ggplot(aes(x = p1, y = p2))+
    geom_tile(aes(fill = efficiency))+
    scale_fill_gradientn(colors = rainbow(7))+
    facet_grid(probability ~ label, labeller = label_parsed)+
    xlab(expression(p[1]))+ 
    ylab(expression(p[2]))+
    theme(
      axis.title.x = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.title.y = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.text.x  = element_text(family = "Helvetica", face = "bold", size = 8),
      axis.text.y  = element_text(family = "Helvetica", face = "bold", size = 8),
      title = element_text(family = "Helvetica", face = "bold", size = 14),
      strip.text.x = element_text(family = "Helvetica", face = "bold", size = 12),
      strip.text.y = element_text(family = "Helvetica", face = "bold", size = 12),
      legend.position = "right",
      legend.title = element_blank(),
      legend.text  = element_text(family = "Helvetica", face = "bold", size = 10)
    )
  
```


## Optimal Allocation to Balance Statistical Efficiency and Ethical Considerations

- Clinical trials are experiments in humans; therefore there is an _**ethical imperative **_ to minimize exposure of study subjects to less successful treatment arms

    + Statistical and ethical considerations may be in conflict
    + An allocation providing a "tradeoff" between statistical efficiency and ethics may be desirable
    
- Consider a 2-arm _**binary outcome**_ trial ($p_E$ and $p_C$ are success rates for Experimental and Control groups, respectively)

- Dual objectives:
    
    i. To maximize study power (Statistical Objective)
    ii. To minimize number of treatment failures in the study (Ethical Objective)
    

##  Design Optimization

- Fixed sample size: $n = n_E + n_C$ ($n_E = \rho n$ and $n_C = (1-\rho)n$)

    + Study goal: to minimize expected number of treatment failures ($ENF = n_Eq_E+n_Cq_C$, where $q_E = 1-p_E$, $q_C = 1-p_C$) while maintaining study power (restricting asymptotic variance of $\widehat{p}_E-\widehat{p}_C$)
    
- Constrained optimization$^*$

$$
\left\{
\begin{array}{c}
  \min\limits_{\rho \in (0, 1)}\left\{n(\rho q_E + (1-\rho)q_C)\right\} \\
  \text{s.t. } \frac{p_Eq_E}{n\rho}+\frac{p_Cq_C}{n(1-\rho)} = C
\end{array}
\right.
$$
($C>0$ is some constant; optimal solution does not depend on $C$)

- Optimal allocation: 
$$
  \rho^* = \frac{\sqrt{p_E}}{\sqrt{p_E}+\sqrt{p_C}}
$$
<br />
<br />
<br />
<br />

<sup>*</sup>Rosenberger WF, Stallard N, Ivanova A, Harper CN, Ricks ML (2001) "_Optimal adaptive designs for binary response trials_", _Biometrics_ **57**, 909-913


## Q & A on Optimal Allocation

<span style="background-color: #6666ff; color:white">
**Q1**:  What is an attractive feature of the allocation $\rho^* = \sqrt{p_E}/(\sqrt{p_E}+\sqrt{p_C})$?.
</span>

<span style="background-color: #6666ff; color:white">
**A1**:  It is always skewed in favor of the more succesfull tratment group (if one exists).
</span>

<span style="background-color: #ffa500; color:white">
**Q2**:  What is a major limitation of the allocation $\rho^* = \sqrt{p_E}/(\sqrt{p_E}+\sqrt{p_C})$?.
</span>

<span style="background-color: #ffa500; color:white">
**A2**:  The model parameters ($p_E$ and $p_C$) are unknown to us! Therefore, this allocation cannot be implemented directly.
</span>


## Example -- Comparison of 3 Allocations

- Two-arm binary outcome trial, $n = 100$

```{r example-3-binary-allocations}
p1 <- seq(0.05, 0.95, by = 0.05)
p2 <- seq(0.05, 0.95, by = 0.05)
n <- 100

allocations <- list(
  equal <- function(pE, pC) {
    list(rho = 0.5, label = "Equal~allocation")
  },
  neymann <- function(pE, pC) {
    list(
      rho = sqrt(pE*(1-pE))/(sqrt(pE*(1-pE)) + sqrt(pC*(1-pC))), 
      label = "Neyman~allocation"
      )
  },
  optimal <- function(pE, pC) {
    list(
      rho = sqrt(pE)/(sqrt(pE) + sqrt(pC)),
      label = "Optimal~allocation"
    )
  }
)

map(allocations, ~ {
allocation_fcn <- .  
crossing(p1, p2) %>% 
  as.list() %>% 
  transpose() %>% 
  map( ~ {
    p <- c(.$p1, .$p2)
    allocation <- allocation_fcn(p[1], p[2])
    tibble::tibble(p1 = p[1], p2 = p[2], rho = allocation$rho, label = allocation$label)
  }) %>% bind_rows()
}) %>% 
  bind_rows() %>% 
  mutate(
    Delta = (p1-p2)/sqrt(p1*(1-p1)/(n*rho)+p2*(1-p2)/(n*(1-rho))),
    Power = 1-pnorm(qnorm(0.975)-Delta)+pnorm(qnorm(0.025)-Delta),
    `ENF/n` = (rho*(1-p1)+(1-rho)*(1-p2))
  ) %>% 
  select(-Delta) %>% 
  gather(key, value, -p1, -p2, -label) %>% 
  ggplot(aes(x = p1, y = p2))+
    geom_tile(aes(fill = value))+
    scale_fill_gradientn(colors = rainbow(7))+
    facet_grid(label ~ key, labeller = label_parsed)+
    xlab(expression(p[1]))+ 
    ylab(expression(p[2]))+
    theme(
      axis.title.x = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.title.y = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.text.x  = element_text(family = "Helvetica", face = "bold", size = 8),
      axis.text.y  = element_text(family = "Helvetica", face = "bold", size = 8),
      title = element_text(family = "Helvetica", face = "bold", size = 14),
      strip.text.x = element_text(family = "Helvetica", face = "bold", size = 12),
      strip.text.y = element_text(family = "Helvetica", face = "bold", size = 12),
      legend.position = "right",
      legend.title = element_blank(),
      legend.text  = element_text(family = "Helvetica", face = "bold", size = 10)
    )
```


## What about multi-arm trial? (1 of 2)

- ($K > 2$)-arm binary outcome trial
- Study goal: to minimize expected treatment failures while restricting the trace of variance-covariance matrix of estimated treatment contrasts

    + Allocation: $\boldsymbol{\rho} = \left(\rho_1, \rho_2, \ldots, \rho_K\right)$ ($n_k = n\rho_k$ -- sample size for the $k$th group)
    
- Constrained optimization problem:
$$
\left\{
\begin{array}{c}
  \min\limits_{\boldsymbol{\rho}}\left\{n\sum_{k = 1}^K{\rho_k q_k}\right\} \\
  \text{s. t. } \sum_{k=2}^K{\frac{p_kq_k}{\rho_k}}+(K-1)\frac{p_1q_1}{\rho_1} = C \\
  \text{and } \sum_{k=1}^K{\rho_k} = 1
\end{array}  
\right.
$$
- Optimal solution ("_Fixed trace optimal allocation_"):

$$
  \rho_1^* = \frac{\sqrt{(K-1)p_1}}{\sqrt{(K-1)p_1}+\sum_{k=2}^K{\sqrt{p_k}}}\text{  and  } \rho_k^* = \frac{\sqrt{p_k}}{\sqrt{(K-1)p_1}+\sum_{k=2}^K{\sqrt{p_k}}}, \quad k = 2, \ldots, K 
$$

## What about multi-arm trial? (2 of 2)

- Binary outcome trial, $K = 3$ treatment groups
- $p_k$=probability of a successful outcome in group $k = 1, 2, 3$

```{r binary-3-trt-fixed-trace-example, echo=FALSE}
prob_tab <- tribble(
  ~id, ~`$p_1$`, ~`$p_2$`, ~`$p_3$`,
  1, 0.10, 0.02, 0.02,
  2, 0.10, 0.30, 0.50,
  3, 0.30, 0.40, 0.60,
  4, 0.40, 0.60, 0.80,
  5, 0.70, 0.30, 0.80,
  6, 0.90, 0.95, 0.99,
  7, 0.35, 0.65, 0.01
)

# calculate Fixed-trace-optimal proportions
prop_tab <- map(seq_len(nrow(prob_tab)), ~ {
  id <- as.numeric(prob_tab[., 1])
  prob_vec <- as.numeric(prob_tab[.,])[-1]
  prop <- round(ftoptimal(prob_vec), 3)
  tribble(
    ~id, ~`$\\rho^*_1$`, ~`$\\rho^*_2$`, ~`$\\rho^*_3$`,
    id, prop[1], prop[2], prop[3])
}) %>% 
  bind_rows()

# join propbabilities and proportions into a single table
inner_join(
  prob_tab, 
  prop_tab,
  by = "id"
) %>% 
  mutate(
    `$\\rho^*_1$`= cell_spec(`$\\rho^*_1$`, "html",
                             color = if_else(id > 2, "red", "black")),
    `$\\rho^*_2$`= cell_spec(`$\\rho^*_2$`, "html", 
                             color = if_else(id > 2, "red", "black")),
    `$\\rho^*_3$`= cell_spec(`$\\rho^*_3$`, "html",
                             color = if_else(id > 2, "red", "black"))
    ) %>% 
  select(-id) %>% 
  knitr::kable("html", escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c("Treatment success probabilities" = 3,
                     "Optimal allocation proportions$^*$" = 3)) %>% 
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(c(1, 3, 5, 7), background = "#99cfe0") %>% 
  row_spec(c(2, 4, 6), background = "#c1e1ec") %>% 
  column_spec(c(1, 2, 3), color = "black")
  #row_spec(-1, background = "#c1e1ec") 
```

<span style="background-color: #6666ff; color:white">
Unlike in the 2-arm case, in the 3-arm case Fixed trace allocation can assign greater proportions to less successful treatments (red rows)
</span>

<br />
<br />
<br />
<br />

<sup>*</sup> $\rho_1^* = \frac{\sqrt{2p_1}}{\sqrt{2p_1}+\sqrt{p_2}+\sqrt{p_2}}$, $\rho_2^* = \frac{\sqrt{p_2}}{\sqrt{2p_1}+\sqrt{p_2}+\sqrt{p_2}}$, $\rho_3^* = \frac{\sqrt{p_3}}{\sqrt{2p_1}+\sqrt{p_2}+\sqrt{p_2}}$

## IV. Optimal Designs for Dose-Response Studies (1 of 2)

- Binary outcome: $Y = \left\{\begin{array}{cl} 1, & \text{if toxicity} \\ 0, & \text{otherwise} \end{array}\right.$
- _**Logistic (2-parameter) dose-response curve**_: 
$$ 
P(d) = \text{Pr}\left(Y=1|d\right)=\frac{1}{1+\exp{\left(-(\alpha+\beta d)\right)}}
$$
- $(\alpha, \beta)$ -- unknown parameters, $\beta > 0$
- $d \in \mathbb{R}$ (after log-transformation); $d_1, \ldots, d_m$ -- distinct dose levels
- $n_k$ obseravtions are made at dose $d_k$; $n = \sum_{k=1}^K{n_k}$ -- total sample size.
- $\rho_k = n_k/n$ -- proportions; $\rho_k > 0$ and $\sum_{k=1}^K{\rho_k} = 1$.
- We are interested in study designs $\xi = \left\{(d_k, \rho_k), \quad k = 1,\ldots, K\right\}$ that are
  + Optimal for estimating $(\alpha, \beta)$ -- entire dose-response curve
  + Optimal for estimating specific percentiles of the DR curve, e.g. $d^{(50)} = -\alpha/\beta$ (dose with 50% toxicity rate).


## IV. Optimal Designs for Dose-Response Studies (2 of 2)

```{r logistic-curve, echo=FALSE}
shinyApp(
  ui = fluidPage(
    sidebarPanel(
      withMathJax(),
      sliderInput("alpha", 
                  "$$\\alpha:$$", 
                  min = -10, 
                  max = 10,
                  step = 0.1,
                  value = 0),
      sliderInput("beta", 
                  "$$\\beta:$$", 
                  min = 0.5, 
                  max = 10,
                  step = 0.5, 
                  value = 1)
    ),
    mainPanel(
      plotOutput("logistic_plot")
    )
  ),
  
  server = function(input, output) {
    output$logistic_plot <- renderPlot({
      d <- seq(-10, 10, by = 0.05)
      
      alpha <- as.numeric(input$alpha)
      beta <- as.numeric(input$beta)
      prob <- 1/(1+exp(-(alpha+beta*d)))
      d_min <- d[1]
      prob_min <- 1/(1+exp(-(alpha+beta*d_min)))
      
      d_max <- d[length(d)]
      prob_max <- 1/(1+exp(-(alpha+beta*d_max)))
      
      d_ave <- (d_min+d_max)/2
      prob_ave <- 1/(1+exp(-(alpha+beta*d_ave)))

      prob_df <- tibble::tibble(d, prob) 
      ggplot()+
          geom_point(data = prob_df, aes(x = d, y = prob), size = 1.5)+
          geom_line(data = prob_df, aes(x = d, y = prob), size = 1.25)+
          # d = d_min
          geom_line(data = tibble::tibble(x = rep(d_min, 2), y = c(0,prob_min)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25)+
          # d = d_ave
          geom_line(data = tibble::tibble(x = rep(d_ave, 2), y = c(0,prob_ave)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25)+
          # d = d_max
          geom_line(data = tibble::tibble(x = rep(d_max, 2), y = c(0,prob_max)), 
                    aes(x = x, y = y), linetype="dashed", size = 1.25)+
          xlab("Dose")+
          ylab("Probability of toxicity")+
          scale_y_continuous(limits = c(0, 1), 
                             breaks = round(c(prob_min, prob_ave, prob_max), 3))+
          scale_x_continuous(breaks = c(d_min, d_ave, d_max),
                             labels = c("minimal", "average", "maximal"))+
    theme(
      axis.title.x = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.title.y = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.text.x  = element_text(family = "Helvetica", face = "bold", size = 8),
      axis.text.y  = element_text(family = "Helvetica", face = "bold", size = 8),
      title = element_text(family = "Helvetica", face = "bold", size = 14),
      strip.text.x = element_text(family = "Helvetica", face = "bold", size = 12),
      strip.text.y = element_text(family = "Helvetica", face = "bold", size = 12),
      legend.position = "right",
      legend.title = element_blank(),
      legend.text  = element_text(family = "Helvetica", face = "bold", size = 10)
    )
    })
  },
  
  options = list(height = 400)
)

```

## Likelihood, MLE, and Fisher Information Matrix (1 of 2)

- _**Likelihood**_: $\mathcal{L}(\alpha, \beta) = \prod\limits_{i = 1}^m{P^{x_i}_i(1-P_i)^{n_i-x_i}}$, where 
  + $x_i = \sum_{j = 1}^{n_i}Y_{ij}$ (total number of toxicities at dose $d_i$)
  + $P_i = P(d_i) = 1/\left\{1+\exp{\left(-(\alpha+\beta d_i)\right)}\right\}$

- $\log{\mathcal{L}(\alpha, \beta)} = \sum_{i = 1}^m\left\{x_i\log{P_i}+(n_i-x_i)\log{\left(1-P_i\right)}\right\}$
- Use the chain rule for differentiation to obtain _**the system of score equations**_:

$$
  \frac{\partial}{\partial{\alpha}}\log\mathcal{L}(\alpha, \beta) = \sum_{i = 1}^m\left\{\frac{x_i}{P_i}-\frac{n_i-x_i}{1-P_i}\right\}\frac{\partial{P_i}}{\partial{\alpha}} = \sum_{i = 1}^m\left(x_i-n_iP_i\right) = 0 \\
    \frac{\partial}{\partial{\beta}}\log\mathcal{L}(\alpha, \beta) = \sum_{i = 1}^m\left\{\frac{x_i}{P_i}-\frac{n_i-x_i}{1-P_i}\right\}\frac{\partial{P_i}}{\partial{\beta}} = \sum_{i = 1}^m\left(x_i-n_iP_i\right)d_i = 0
$$

- Solve the score equations (numerically) to obtain $(\widehat{\alpha}, \widehat{\beta})$ -- the MLE of $(\alpha, \beta)$

## Likelihood, MLE, and Fisher Information Matrix (2 of 2)

- _**Fisher Information Matrix**_ $\boldsymbol{M}(\alpha, \beta) = -\mathbf{E}\left(\frac{\partial^2}{\partial{\boldsymbol{\theta}}\partial{\boldsymbol{\theta}'}}\log\mathcal{L}(\boldsymbol{\theta})\right)$, where $\boldsymbol{\theta} = (\alpha, \beta)$
- After doing some algebra, we obtain:

$$
\left(
\begin{array}{cc}
  \sum_{i=1}^m\rho_i\frac{e^{-a_i}}{(1+e^{-a_i})^2} & 
  \sum_{i=1}^m\rho_i\frac{e^{-a_i}}{(1+e^{-a_i})^2}d_i \\
  \sum_{i=1}^m\rho_i\frac{e^{-a_i}}{(1+e^{-a_i})^2}d_i & 
  \sum_{i=1}^m\rho_i\frac{e^{-a_i}}{(1+e^{-a_i})^2}d^2_i
\end{array}
\right),
$$

where $\rho_i = n_i/n$ -- allocation proportion at dose $d_i$ and $a_i = \alpha+\beta d_i$

- We are interested in finding:

    + _**D-optimal design**_ that minimizes $\text{det}\boldsymbol{M}^{-1}(\alpha, \beta)$
    + _**A-optimal design**_ that minimizes $\text{trace}\boldsymbol{M}^{-1}(\alpha, \beta)$
    

## D-optimal Design$^*$

_**Result 1**_: The design $\left\{(d_1^*, 0.5), (d_2^*, 0.5)\right\}$, where $d_1^*$ and $d_2^*$ satisfy: 
$$
\begin{array}{c}
  \alpha + \beta d_1^* = -C_D \\
    \alpha + \beta d_2^* = C_D
\end{array}
$$
is D-optimal.

- $C_D = 1.5434$ -- it is obtained by maximizing the function $\frac{c^2e^{2c}}{(1+e^c)^4}$
```{r logistic-curve-and-d-optimal-design, echo=FALSE}
shinyApp(
  ui = fluidPage(
    sidebarPanel(
      withMathJax(),
      sliderInput("alpha", 
                  "$$\\alpha:$$", 
                  min = -10, 
                  max = 10,
                  step = 0.1,
                  value = 0),
      sliderInput("beta", 
                  "$$\\beta:$$", 
                  min = 0.5, 
                  max = 10,
                  step = 0.5, 
                  value = 1)
    ),
    mainPanel(
      plotOutput("logistic_plot")
    )
  ),
  
  server = function(input, output) {
    output$logistic_plot <- renderPlot({
      d <- seq(-10, 10, by = 0.05)
      
      alpha <- as.numeric(input$alpha)
      beta <- as.numeric(input$beta)
      prob <- 1/(1+exp(-(alpha+beta*d)))
      xi_dopt <- list(
        x = (c(-1.5434, 1.5434)-alpha)/beta,
        w = c(0.5, 0.5)
      )
  model <- tibble::tibble(
    x = d, 
    y = prob
  )
  design <- tibble::tibble(x = xi_dopt$x, w = xi_dopt$w)                    

  ggplot()+
    geom_line(data = model, aes(x = x, y = y), size=1.5)+
    xlab("Dose")+
    ylab("Probability")+
    geom_col(data=design, aes(x = x, y = w*1), width = 0.25, fill = "red", alpha = 0.5)+
    scale_x_continuous(breaks = round(c(xi_dopt$x), 2))+
    scale_y_continuous(sec.axis = sec_axis(~./1, name = "D-optimal allocation proportion", breaks = round(xi_dopt$w, 2)))+
    theme(
      axis.title.x = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.title.y = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.text.x  = element_text(family = "Helvetica", face = "bold", size = 8),
      axis.text.y  = element_text(family = "Helvetica", face = "bold", size = 8),
      title = element_text(family = "Helvetica", face = "bold", size = 14),
      strip.text.x = element_text(family = "Helvetica", face = "bold", size = 12),
      strip.text.y = element_text(family = "Helvetica", face = "bold", size = 12),
      legend.position = "right",
      legend.title = element_blank(),
      legend.text  = element_text(family = "Helvetica", face = "bold", size = 10)
    )
    })
  },
  
  options = list(height = 400)
)

```

<br />
<br />
<br />
<br />

<sup>*</sup>Mathew T, Sinha BK (2001) "_Optimal designs for binary data under logistic regression_", _Journal of Statistical Planning and Inference_ **93**, 295-307


## A-optimal Design$^*$

- There is no complete solution to the A-optimal problem!
- Within the class of _**symmetric designs**_, the A-optimal design is given by $\left\{(c, 0.5), (-c, 0.5)\right\}$, where $c$ is obtained by minimizing the function

$$
  \frac{\alpha^2+\beta^2}{c^2e^c/(1+e^c)^2}+ \frac{1}{e^c/(1+e^c)^2}
$$

- Within the class of _**two-point designs**_ $\left\{(c_1, \rho), (c_2, 1-\rho)\right\}$, the A-optimal design is found numerically, by minimizing w.r.t. $c_1, c_2$, and $\rho$ the function
$$
  \frac{\rho C_1(\beta^2+(c_1-\alpha)^2)+(1-\rho)C_2(\beta + (c_2-\alpha)^2)}{(\rho C_1 + (1-\rho)C_2)(\rho c_1^2C_1+(1-\rho)c_2^2C_2) - (\rho c_1C_1 + (1-\rho)c_2C_2)^2},
$$
where $C_i = \frac{e^{-c_i}}{(1+e^{-c_i})^2}, \quad i = 1, 2$.

- Thus, in the latter case, both the A-optimal dose-levels and allocation proportions depend on $(\alpha, \beta)$

<br />
<br />
<br />
<br />

<sup>*</sup>Mathew T, Sinha BK (2001) "_Optimal designs for binary data under logistic regression_", _Journal of Statistical Planning and Inference_ **93**, 295-307


## Examples -- A-optimal Designs

- $(\alpha, \beta) = (10, 5)$
- _**two-point design**_: $\left\{(-2.38, 0.59), (2.38, 0.41)\right\}$, $\text{trace}\boldsymbol{M}(\alpha, \beta) = 287.29$

- _**two-point symmetric design**_: $\left\{(-2.33, 0.5), (2.33, 0.5)\right\}$, $\text{trace}\boldsymbol{M}(\alpha, \beta) = 297.31$

```{r example-logistic-a-optimal}

d <- seq(-4, 3, by = 0.05)
      
alpha <-10
beta <- 5

prob <- 1/(1+exp(-(alpha+beta*d)))
xi1_aopt <- list(
  x = c(-2.33, 2.33),
  w = c(0.5, 0.5)
)

xi2_aopt <- list(
  x = c(-2.38, 2.38),
  w = c(0.5944, 0.4056)
)
      
model <- bind_rows(
  tibble::tibble(x = d, y = prob, class = "two-point symmetric"),
  tibble::tibble(x = d, y = prob, class = "two-point")
)

design <- bind_rows(
  tibble::tibble(x = xi1_aopt$x, w = xi1_aopt$w, class = "two-point symmetric"),
  tibble::tibble(x = xi2_aopt$x, w = xi2_aopt$w, class = "two-point")
)

ggplot()+
    geom_line(data = model, aes(x = x, y = y), size=1.5)+
    xlab("Dose")+
    ylab("Probability")+
    geom_col(data=design, aes(x = x, y = w*1), width = 0.25, fill = "red", alpha = 0.5)+
    scale_x_continuous(breaks = c(-3, -2, 2, 3))+
    scale_y_continuous(sec.axis = sec_axis(~./1, name = "D-optimal allocation proportion", breaks = sort(c(round(xi1_aopt$w, 2), round(xi2_aopt$w, 2)))))+
  facet_grid(~ class)+
    theme(
      axis.title.x = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.title.y = element_text(family = "Helvetica", face = "bold", size = 12),
      axis.text.x  = element_text(family = "Helvetica", face = "bold", size = 8),
      axis.text.y  = element_text(family = "Helvetica", face = "bold", size = 8),
      title = element_text(family = "Helvetica", face = "bold", size = 14),
      strip.text.x = element_text(family = "Helvetica", face = "bold", size = 12),
      strip.text.y = element_text(family = "Helvetica", face = "bold", size = 12),
      legend.position = "right",
      legend.title = element_blank(),
      legend.text  = element_text(family = "Helvetica", face = "bold", size = 10)
    )

```

## Homework for Lecture #2 (1 of 4)

_**Problem 1**_: Consider a 2-arm trial with normal outcomes and equal variances (when balanced allocation $(0.5, 0.5)$ is optimal). Let $\delta = \frac{\mu_E-\mu_C}{\sigma}$ denote the standardized mean difference, a.k.a. "effect size"

(1) Write down an analytical expression for _**Power**_ in terms of $\delta, n, \rho, z_{1-\alpha}$
(2) Consider 3 allocation strategies: $1:1$ ($\rho = 0.5$); $2:1$ ($\rho = 2/3$); $3:1$ ($\rho = 3/4$). Assuming effect size $\delta = 0.5$ and significance level $\alpha = 0.05$, plot the _**Power**_ curve for each of the three allocation strategies for $n = 25, 26, \ldots, 199, 200$.
(3) From (2), what sample size is requiered to achive 80% (90%) power? Fill out the table: 

```{r problem1-table, echo=FALSE}
tab <- tibble::tibble(
  `Allocation` = c('1:1', '2:1', '3:1'),
  `Sample size for 80% power` = c('', '', ''),
  `Sample size for 90% power` = c('', '', '')
)
knitr::kable(tab, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(c(1,3), background = "#99cfe0") %>% 
  row_spec(2, background = "#c1e1ec")
```

## Homework for Lecture #2 (2 of 4)

_**Problem 2**_: Consider a 2-arm trial with normal outcomes and unequal variances. Suppose the study goal is to estimate treatment difference $(\mu_E-\mu_C)$. In this case, the D<sub>A</sub>-allocation is Neyman allocation: $(\rho^*, 1-\rho^*)$, where $\rho^* = \sigma_E/(\sigma_E+\sigma_C)$.

(1) Show that the relative efficiency (*RE*) of equal alocation $(0.5, 0.5)$ vs. D<sub>A</sub>-optimal allocation is $RE = \frac{\left(\sigma_E+\sigma_C\right)^2}{2(\sigma_E^2+\sigma_C^2)}$
(2) Show that if $\sigma_E = 0.5\sigma_C$ or $\sigma_E = 2\sigma_C$, then $RE = 0.9$
(3) Show that if $\frac{\sigma_E}{\sigma_C}\rightarrow 0$ or $\frac{\sigma_C}{\sigma_E}\rightarrow +\infty$ (one of the variances is much larger than the other one), then $RE\rightarrow 0.5$


## Homework for Lecture #2 (3 of 4)

_**Problem 3**_: Consider a 2-arm trial with binary outcomes, where success rates for the experimental and control groups are $p_E$ and $p_C$ respectively. The Neyman allocation is $(\rho^*, 1-\rho^*)$, where 
$$\rho^* = \frac{\sqrt{p_E(1-p_E)}}{\sqrt{p_E(1-p_E)}+\sqrt{p_C(1-p_C)}}.$$

(1) Show that Neyman allocation assigns greater proportion of patients to an inferior (less successful) treatment arm, if $p_E + p_C > 1$
(2) Under which two conditions is Neyman allocation the same as the equal allocation (0.5, 0.5)? 

## Homework for Lecture #2 (4 of 4)

_**Problem 4**_: Consider a toxicology study where the goal is to estimate the dose-toxicity relationship, assumed to follow a 2-parameter logistic model: $\text{Pr}\left(Y=1|d\right) = 1/\left\{1+\exp{\left(-(\alpha+\beta d)\right)}\right\}$. A pilot study was conducted and the results were as follows:

```{r problem4-table, echo=FALSE}
alpha <- 1
beta <- 1

d <- c(-1.5, 0, 1.5)
logistic_fcn <- function(dose, a, b) {1/(1+exp(-a-b*dose))}

dose <- sample(d, 10, TRUE)
prob <- logistic_fcn(dose, alpha, beta)
n <- map_dbl(d, ~ sum(as.numeric(. == dose)))
Y <- rbinom(length(dose), 1, prob) 
x <- map_dbl(d, ~ sum(Y[. == dose]))
  
d <- c(-1.5, 0, 1.5)
n <- c(4, 2, 4)
x <- c(1, 1, 4)
  
tab <- tibble::tibble(
  `$d_i$` = d,
  `$n_i$` = n,
  `$x_i$` = x
)
knitr::kable(tab, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  row_spec(0, bold = T, color = "white", background = "#6666ff") %>% 
  row_spec(c(1,3), background = "#99cfe0") %>% 
  row_spec(2, background = "#c1e1ec")
```

(1) Fit a logistic model to the data, obtain the MLE $(\widehat{\alpha}, \widehat{\beta})$ of $(\alpha, \beta)$ and its variance-covariance matrix.
(2) Suppose we want to design a new study. Based on the results from (1), what should be D-optimal and A-optimal designs?


## Backup slides


## FIM derivation for 2-parameter logistic model

- _**Likelihood**_: $\mathcal{L}(\alpha, \beta) = \prod\limits_{i=1}^m{P_i^{x_i}(1-P_i)^{n_i-x_i}}$, where

    + $x_i = \sum_{i = 1}^{n_i}Y_{ij}$ (total number of toxicities at dose $d_i$)
    + $P_i = P(d_i) = 1/\left\{1+\exp{\left(-(\alpha+\beta d_i)\right)}\right\}$

- $\log{\mathcal{L}(\alpha, \beta)} = \sum_{i = 1}^m\left\{x_i\log{P_i}+(n_i-x_i)\log{(1-P_i)}\right\}$
- $\frac{\partial}{\partial{\alpha}}\log{\mathcal{L}(\alpha, \beta)} = \sum_{i = 1}^m\left\{\frac{x_i}{P_i}-\frac{n_i-x_i}{1-P_i}\right\}\frac{\partial{P_i}}{\partial{\alpha}} = \left[\frac{\partial{P_i}}{\partial{\alpha}} = P_i(1-P_i)\right] = \sum_{i = 1}^m{(x_i-n_iP_i)}$
- $\frac{\partial}{\partial{\beta}}\log{\mathcal{L}(\alpha, \beta)} = \sum_{i = 1}^m\left\{\frac{x_i}{P_i}-\frac{n_i-x_i}{1-P_i}\right\}\frac{\partial{P_i}}{\partial{\beta}} = \left[\frac{\partial{P_i}}{\partial{\beta}} = P_i(1-P_i)d_i\right] = \sum_{i = 1}^m{(x_i-n_iP_i)d_i}$
- $\frac{\partial^2}{\partial{\alpha^2}}\log{\mathcal{L}(\alpha, \beta)} = \frac{\partial}{\partial{\alpha}}\sum_{i = 1}^m{(x_i-n_iP_i)} = -\sum_{i = 1}^m{n_iP_i(1-P_i)}$
- $\frac{\partial^2}{\partial{\alpha}\partial{\beta}}\log{\mathcal{L}(\alpha, \beta)} = \frac{\partial}{\partial{\beta}}\sum_{i = 1}^m{(x_i-n_iP_i)} = -\sum_{i = 1}^m{n_iP_i(1-P_i)d_i}$
- $\frac{\partial^2}{\partial{\beta^2}}\log{\mathcal{L}(\alpha, \beta)} = \frac{\partial}{\partial{\beta}}\sum_{i = 1}^m{(x_i-n_iP_i)d-i} = -\sum_{i = 1}^m{n_iP_i(1-P_i)d_i^2}$

